% ============================================================
% appendices/A_phi_minus_one_third.tex  (FULL REPLACEMENT)
% ============================================================

\section{Self-Consistent Fixed-Point Derivation of the
\texorpdfstring{$\boldsymbol{\Phi^{-1/3}}$}{Phi(-1/3) Scaling}}
\label{app:phi_minus_one_third}

This appendix provides an explicit self-consistent fixed-point
\emph{construction} of the cubic-root scaling
$\delta t_{\min}\propto \Phi^{-1/3}$ for photon-limited temporal
localization of a diffusive particle.
The derivation is purely operational and does not constitute
decision-based inference evidence.
No microscopic assumptions are invoked beyond normal diffusion,
Poisson detection statistics, finite measurement noise, and the
existence of a self-consistent closure.

\subsection{Setup and assumptions}

We consider one-dimensional Brownian motion with diffusion coefficient
$D$,
\begin{equation}
\langle x^2(t)\rangle = 2Dt,
\end{equation}
observed through a Poisson point process of detection events with
constant photon flux $\Phi$.
Each detection yields an independent position estimate with Gaussian
point-spread uncertainty $\sigma_m$.

The assumptions are:
\begin{enumerate}
\item Normal diffusion with finite $D$ and no drift.
\item Poissonian detection statistics with rate $\Phi$.
\item Independent measurement noise with variance $\sigma_m^2$ per
detection.
\item A fixed operational distinguishability threshold $D^\ast$
(Appendix~\ref{app:fisher_qfi}) used as an internal decision criterion
within the construction.
\end{enumerate}

\subsection{Operational localization variance}

Let $\sigma^2(t)$ denote the effective localization variance conditioned
on the measurement record accumulated up to time $t$.
Two competing mechanisms determine $\sigma^2(t)$:
\begin{itemize}
\item \textbf{Diffusive spreading:} uncertainty increases as $2Dt$.
\item \textbf{Measurement-induced reduction:} information accumulates at
rate $\Phi$, reducing variance as $\sigma_m^2/(\Phi t)$.
\end{itemize}

To leading operational order, the localization variance is therefore
\begin{equation}
\sigma^2(t)\;\sim\; 2Dt \;+\; \frac{\sigma_m^2}{\Phi t}.
\label{eq:sigma_balance}
\end{equation}
This expression captures the feedback between stochastic spreading and
finite-rate information acquisition under the specified observation
model.

\subsection{Self-consistent fixed point}

The characteristic time scale associated with the construction is
obtained by minimizing $\sigma^2(t)$ with respect to $t$.
Differentiating Eq.~\eqref{eq:sigma_balance} gives
\begin{equation}
\frac{d\sigma^2}{dt}
= 2D - \frac{\sigma_m^2}{\Phi t^2}.
\end{equation}
Setting the derivative to zero yields the fixed point
\begin{equation}
t_\ast
\;\sim\;
\left(\frac{\sigma_m^2}{D\,\Phi}\right)^{1/3}.
\label{eq:t_star}
\end{equation}
This time scale marks the balance between diffusion-induced uncertainty
growth and measurement-induced uncertainty reduction.

\subsection{Temporal distinguishability}

Temporal distinguishability arises because the diffusive state depends on
time.
For small temporal offsets $\delta t$, the Fisher information about time
scales inversely with the localization variance,
\begin{equation}
I_T(t) \;\sim\; \frac{t}{\sigma^2(t)},
\label{eq:fi_scaling}
\end{equation}
up to dimensionless constants that do not affect scaling.
This relation expresses how changes in diffusive width encode temporal
information under the assumed channel.

Evaluating Eq.~\eqref{eq:fi_scaling} at the self-consistent scale $t_\ast$
yields
\begin{equation}
I_T(t_\ast)
\;\sim\;
\frac{t_\ast}{\sigma^2(t_\ast)}
\;\sim\;
\frac{\Phi^{2/3}}{D^{1/3}\sigma_m^{4/3}},
\end{equation}
where numerical prefactors have been suppressed.

\subsection{Operational time resolution}

The minimal resolvable temporal offset $\delta t_{\min}$ follows from the
operational decision criterion
\begin{equation}
\delta t^2\, I_T \;\gtrsim\; 2D^\ast,
\label{eq:decision_again}
\end{equation}
which is used here as an internal consistency condition.
Substituting the scaling of $I_T(t_\ast)$ gives
\begin{equation}
\delta t_{\min}
\;\sim\;
\left(\frac{\sigma_m^2}{D^2\,\Phi}\right)^{1/3},
\end{equation}
or equivalently,
\begin{equation}
\delta t_{\min}\propto \Phi^{-1/3}.
\end{equation}

\subsection{Interpretation}

The cubic-root scaling does not arise from dimensional analysis alone.
It emerges from a self-consistent balance between:
\begin{enumerate}
\item stochastic diffusive spreading,
\item Poisson-limited information acquisition, and
\item a fixed operational distinguishability threshold.
\end{enumerate}
Changing any of these ingredients modifies the resulting scaling.

\subsection{Epistemic status and validity}

The $\Phi^{-1/3}$ law obtained here is epistemic and
construction-based.
It reflects limitations imposed by the specified observation channel
and vanishes in the limits $\Phi\to\infty$ or $\sigma_m\to 0$, indicating
improved accessibility rather than intrinsic discreteness of time.
Its validity is restricted to the self-consistent regime defined by the
assumptions above; pre-asymptotic effects or alternative observation
channels modify the effective behavior.

\subsection{Connection to generalizations}

For anomalous diffusion with $\mathrm{MSD}\sim t^\alpha$, the same
self-consistent closure logic yields
$\delta t_{\min}\propto\Phi^{-1/(2+\alpha)}$, as derived in
Appendix~B.
The normal diffusion case corresponds to $\alpha=1$ and serves as the
canonical baseline of the construction.

% ============================================================
% End of appendices/A_phi_minus_one_third.tex
% ============================================================
