% ============================================================
% appendices/B_ctrw_derivations.tex  (FULL REPLACEMENT)
% ============================================================

\section{Anomalous Diffusion: CTRW Fixed-Point Construction}
\label{app:ctrw_derivations}

This appendix generalizes the self-consistent fixed-point construction
of Appendix~\ref{app:phi_minus_one_third} to anomalous diffusion
described by continuous-time random walks (CTRW).
The purpose is to construct the photon-limited scaling
\[
\delta t_{\min}\propto \Phi^{-1/(2+\alpha)}
\]
under explicit, operational assumptions.
No claim of decision-based inference or data-driven exponent extraction
is made; the transport exponent $\alpha$ is treated as an imposed
property of the latent process.

\subsection{CTRW model and transport exponent}

We consider a CTRW characterized by independent spatial jumps with finite
second moment and waiting times drawn from a heavy-tailed distribution.
In the long-time limit, the mean-square displacement (MSD) scales as
\begin{equation}
\langle x^2(t)\rangle \;\sim\; C\, t^\alpha,
\qquad 0<\alpha<2,
\end{equation}
where $C$ is a generalized transport coefficient.
The regimes $\alpha<1$, $\alpha=1$, and $\alpha>1$ correspond to
subdiffusion, normal diffusion, and superdiffusion, respectively.

\paragraph{Assumptions.}
The construction relies on:
\begin{enumerate}
\item A prescribed asymptotic MSD exponent $\alpha$ characterizing the
latent transport process.
\item Poissonian detection statistics with constant photon flux $\Phi$.
\item Independent measurement noise with variance $\sigma_m^2$ per
detection.
\item A fixed operational distinguishability threshold $D^\ast$
(Appendix~\ref{app:fisher_qfi}) used as an internal consistency criterion.
\end{enumerate}

\subsection{Operational localization balance}

As in Appendix~\ref{app:phi_minus_one_third}, the effective localization
variance $\sigma^2(t)$ reflects a balance between anomalous spreading and
measurement-induced localization.
Operationally,
\begin{equation}
\sigma^2(t)\;\sim\; C\, t^\alpha \;+\; \frac{\sigma_m^2}{\Phi t}.
\label{eq:ctrw_balance}
\end{equation}
The first term captures the imposed anomalous transport of the latent
trajectory, while the second encodes finite-rate information acquisition
through Poisson-limited measurements.

\subsection{Self-consistent fixed-point time scale}

The characteristic time scale associated with the construction is
obtained by minimizing $\sigma^2(t)$ with respect to $t$.
Differentiating Eq.~\eqref{eq:ctrw_balance} gives
\begin{equation}
\frac{d\sigma^2}{dt}
= \alpha C t^{\alpha-1}
- \frac{\sigma_m^2}{\Phi t^2}.
\end{equation}
Setting the derivative to zero yields the fixed point
\begin{equation}
t_\ast \;\sim\;
\left(\frac{\sigma_m^2}{C\,\Phi}\right)^{\!1/(2+\alpha)}.
\label{eq:ctrw_t_star}
\end{equation}
This scale marks the balance between anomalous spreading and finite-rate
measurement-induced uncertainty reduction.

\subsection{Temporal distinguishability}

Temporal distinguishability arises because the width of the anomalous
diffusion process depends on time.
For small temporal offsets $\delta t$, the Fisher information about time
scales inversely with the localization variance,
\begin{equation}
I_T(t) \;\sim\; \frac{t}{\sigma^2(t)},
\label{eq:ctrw_fi_scaling}
\end{equation}
up to dimensionless constants.
This relation expresses how changes in the MSD encode temporal
information under the specified observation channel.

Evaluating Eq.~\eqref{eq:ctrw_fi_scaling} at the self-consistent scale
$t_\ast$ gives
\begin{equation}
I_T(t_\ast)
\;\sim\;
\Phi^{\frac{2}{2+\alpha}}\,
C^{-\frac{1}{2+\alpha}}\,
\sigma_m^{-\frac{2\alpha}{2+\alpha}},
\end{equation}
where numerical prefactors have been suppressed, as they do not affect
scaling exponents.

\subsection{Operational time-resolution scaling}

The minimal resolvable temporal offset $\delta t_{\min}$ follows from
the operational decision criterion
\begin{equation}
\delta t^2\, I_T \;\gtrsim\; 2D^\ast,
\end{equation}
used here as an internal closure condition.
Substituting the scaling of $I_T(t_\ast)$ yields
\begin{equation}
\delta t_{\min}
\;\sim\;
\Phi^{-\,\frac{1}{2+\alpha}},
\label{eq:ctrw_scaling}
\end{equation}
up to prefactors depending on $C$, $\sigma_m$, and $D^\ast$.

\subsection{Consistency with normal diffusion}

Setting $\alpha=1$ in Eq.~\eqref{eq:ctrw_scaling} recovers the cubic-root
law $\delta t_{\min}\propto \Phi^{-1/3}$ obtained in
Appendix~\ref{app:phi_minus_one_third}.
The CTRW result therefore represents a continuous generalization of the
same fixed-point construction rather than a distinct inference
mechanism.

\subsection{Numerical consistency checks}

Monte Carlo simulations of CTRW trajectories with prescribed $\alpha$
are used as a consistency check of the fixed-point construction.
An $\alpha$-sweep comparing extracted slopes against the imposed
theoretical exponent $-1/(2+\alpha)$ is included in the frozen Evidence
Pack (Figure~9).
These simulations do not constitute discovery or validation of the
scaling law; they verify that the numerical implementation reproduces
the behavior implied by the assumed transport exponent and observation
model.

\subsection{Epistemic status and regime of validity}

The CTRW scaling obtained here is epistemic and construction-based.
It reflects limitations imposed by anomalous transport combined with a
Poisson-limited observation channel and vanishes in the limit of
unbounded information rate.
Deviations from Eq.~\eqref{eq:ctrw_scaling} diagnose pre-asymptotic
regimes, breakdown of the assumed MSD scaling, or unmodeled information
channels rather than ontological constraints.

\subsection{Falsifiability}

Observation of sustained scaling steeper than
$\Phi^{-1/(2+\alpha)}$ under the same CTRW and observation assumptions
would falsify the rate-limited hypothesis for anomalous diffusion.
Conversely, systematic deviations toward weaker scaling indicate noise,
finite-sample effects, or model misspecification.

% ============================================================
% End of appendices/B_ctrw_derivations.tex
% ============================================================
