% ============================================================
% appendices/E_reproducibility_checklist.tex  (FULL REPLACEMENT)
% ============================================================

\section{Reproducibility Checklist}
\label{app:reproducibility}

This appendix provides a concise and auditable checklist documenting the
reproducibility status of all computational results reported in the paper.
Its sole purpose is verification: to enable independent reproduction of all
numerical results and figures directly from the public repository, without
additional correspondence, clarification, or author intervention.

\subsection{Repository state}

\begin{itemize}
\item All reported results correspond to a single, immutable, tagged repository
state. The tag used for this manuscript is \texttt{drt-paper-freeze-2026-01-14}.
\item The Evidence Pack (all simulation outputs and generated figures) is frozen
at this tag and not modified during manuscript preparation.
\item All \LaTeX{} sources, simulation scripts, analysis code, build
configuration, and auxiliary files required for reproduction are contained
entirely within the repository.
\item No external data sources, unpublished datasets, or private dependencies
are required.
\end{itemize}

\subsection{Build environment}

\begin{itemize}
\item The numerical pipeline is implemented in Python using explicitly specified
dependencies listed in the repository requirements file.
\item Figure generation relies exclusively on precomputed simulation outputs
stored as structured JSON artifacts.
\item No figure is generated from live or on-the-fly simulation output during
manuscript compilation.
\item No proprietary software, closed-source libraries, or
platform-specific tools are required.
\end{itemize}

\subsection{One-command reproducibility}

From a clean environment, all numerical results, figures, and the manuscript PDF
can be reproduced via the command sequence:
\begin{verbatim}
make setup
make sims
make figs
make pdf
\end{verbatim}
This sequence installs all dependencies, executes all simulations, generates all
figures, and compiles the manuscript without manual intervention or interactive
input.

\subsection{Randomness and determinism}

\begin{itemize}
\item All stochastic simulations employ explicit random seeds recorded in their
corresponding output files.
\item Re-running the full pipeline with identical inputs reproduces all figures
and numerical results within floating-point tolerance.
\item Minor platform-dependent roundoff effects or \LaTeX{} nondeterminism do not
affect extracted scaling exponents, operational decision thresholds, or reported
conclusions.
\item No interactive choices, hidden parameters, manual overrides, or
post-processing steps influence any reported result.
\end{itemize}

\subsection{Traceability and auditability}

\begin{itemize}
\item Each figure included in the manuscript corresponds to a uniquely
identified file in the \texttt{figures/} directory.
\item Each figure is traceable to specific raw simulation outputs stored in the
\texttt{results/} directory.
\item The mapping between analytic claims, simulation scripts, output artifacts,
and manuscript sections is documented explicitly in the project
\texttt{RUNBOOK}.
\item Any modification of simulation parameters, analysis logic, or figure
generation scripts would produce a detectable change in the repository state.
\end{itemize}

\subsection{Scope of reproducibility}

The following elements are fully reproducible from the tagged repository state:

\begin{itemize}
\item All numerical simulations and extracted scaling exponents.
\item All figures and tables included in the manuscript.
\item The compiled manuscript PDF corresponding to the tagged release.
\end{itemize}

The following elements are explicitly \emph{not} claims of reproducibility:

\begin{itemize}
\item Conceptual interpretation of results.
\item Ontological classification of bounds.
\item Historical motivation or narrative framing.
\end{itemize}

\subsection{Integrity statement}

No reported numerical result is obtained through manual tuning, selective
reporting, adaptive parameter choice, or post-hoc adjustment.
All figures and quantitative values reflect the direct output of the documented,
deterministic computational pipeline.
This checklist is intended to render any deviation from the reported results
immediately detectable by independent inspection.

% ============================================================
% End of appendices/E_reproducibility_checklist.tex
% ============================================================
